HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-15 scnn.tex
method: efficient convolution

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\L}{\b{L}}
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\x}{\b{x}}
\newcommand{\bu}{\b{u}}
\newcommand{\f}{\b{f}}
\newcommand{\trans}{^\intercal}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

As there is no notion of translation on a graph, we cannot really convolve two graph signals. We can, however, convolve a signal with a kernel defined in the spectral domain. More precisely, we are filtering a graph signal by a kernel. Given the convolution kernel
$h: \R_+ \rightarrow \R$, a signal $\f$ is filtered as
\begin{equation*} \label{eqn:graph_convolution}
	h(L) \f = \U h(\bLambda) \U\trans \f,
\end{equation*}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii} = h(\bLambda_{ii})$, and $h(\bLambda) \hat{\f}$ is the spectral representation of $\f$ after filtering.

Contrary to classical signal processing on Euclidean domains, the kernel $h$ has no single representation in the vertex domain and cannot be translated on the graph. It can however be localized on any vertex $v_i$ by the convolution with a Kronecker delta $\b \delta_i \in \R^{N_{pix}}$,\footnote{A Kronecker delta is the signal $\b \delta_i \in \R^{N_{pix}}$ that is zero everywhere except on vertex $v_i$ where it takes the value one.} that is $h(\L) \b \delta_i$, a column of $h(\L)$.
That is useful to visualize kernels.
If the graph is not regular, i.e., all nodes don't have the same number of neighbors, and all distances are not equal, the effect of the kernel will slightly differ from one node to another. While there is no perfect sampling of the sphere, these differences are negligible as the structure of the whole graph is very regular. However, when considering only parts of the sphere, one can observe important border effects (see \figref{border_effects} and~\ref{sec:border_effects}).
Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with all localizations of the kernel:
\begin{equation*} \label{eqn:graph_convolution}
	(h(L) \f)_i = \langle h(L) \b \delta_i, \f \rangle.
\end{equation*}

\paragraph{Example}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
	\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}: \R_+ \rightarrow \R^{N_{pix}}$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation*}
	\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation*}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(\lambda)=e^{-\tau t \lambda}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$, which are a generalization of squared frequencies. Hence, the kernel $K_t$ can also be considered as a generalization of the Gaussian kernel on the sphere. \figref{gaussian_filters_comparizon} shows the effect of the convolution by diffusing a unit of heat for $\tau=1$ at various times $t$. By comparing the graph convolution with the spherical symmetric Gaussian smoothing, we observe that both techniques lead to similar results (see \figref{gaussian_filters_comparizon}). While the graph convolution is different from a true spherical convolution, it can, providing the correct parameterization, well approximate it. \figref{gaussian_filters_visualization} provides additional visualizations of the convolution kernel $K_t$.

@@

This result is another sign that the constructed graph is able to capture the spherical
structure of the HEALPix sampling. In some applications, where the
exactitude of the convolution is not a requirement, such as de-noising, graph convolution could be used instead of using the spherical harmonics.
Depending on the convolution kernel, this may reveal much more efficient.
\todo{Show the figure comparing speed of low-pass filtering? It's 10x faster.}
The fact that both convolutions are not identical is not relevant in a neural network setting, as the convolution kernels are learned anyway.
% not ideal as the graph convolution depends on the localization

\subsection{Efficient convolutions}
\label{sec:efficient_convolution}
\assign{Michaël}

Transition from spectral to spatial domain.

With the Fourier basis. Because we don't have an FFT, that costs $O(n^3)$ for the eigendecomposition, plus $O(n^2)$ for the transform (to be done for each forward and backward pass).
\begin{itemize}
	\item We need efficient convolution, hence the Chebysheff trick
	\item Explain the different between a polynomial filter convolution and a traditional patch restricted convolution
	\item Define spherical CNN using graph CNN
\end{itemize}

>>>>

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\L}{\b{L}}
\newcommand{\tL}{\tilde{\L}}
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\x}{\b{x}}
\newcommand{\bu}{\b{u}}
\newcommand{\f}{\b{f}}
\newcommand{\trans}{^\intercal}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\bO}{\mathcal{O}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

As there is no notion of translation on a graph, we cannot really convolve two graph signals. We can, however, convolve a signal with a kernel defined in the spectral domain. More precisely, we are filtering a graph signal by a kernel. Given the convolution kernel
$h: \R_+ \rightarrow \R$, a signal $\f \in \R^{N_{pix}}$ on the graph is filtered as
\begin{equation} \label{eqn:graph_convolution}
	h(L) \f = \U h(\bLambda) \U\trans \f,
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii} = h(\bLambda_{ii})$, and $h(\bLambda) \hat{\f}$ is the spectral representation of $\f$ after filtering.

Contrary to classical signal processing on Euclidean domains, the kernel $h$ has no single representation in the vertex domain and cannot be translated on the graph. It can however be localized on any vertex $v_i$ by the convolution with a Kronecker delta $\b \delta_i \in \R^{N_{pix}}$,\footnote{A Kronecker delta is the signal $\b \delta_i \in \R^{N_{pix}}$ that is zero everywhere except on vertex $v_i$ where it takes the value one.} that is $h(\L) \b \delta_i$, a column of $h(\L)$.
That is useful to visualize kernels.
If the graph is not regular, i.e., all nodes don't have the same number of neighbors, and all distances are not equal, the effect of the kernel will slightly differ from one node to another. While there is no perfect sampling of the sphere, these differences are negligible as the structure of the whole graph is very regular. However, when considering only parts of the sphere, one can observe important border effects (see \figref{border_effects} and~\ref{sec:border_effects}).
Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with all localizations of the kernel:
\begin{equation*}
	(h(L) \f)_i = \langle h(L) \b \delta_i, \f \rangle.
\end{equation*}

\paragraph{Example}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
	\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}: \R_+ \rightarrow \R^{N_{pix}}$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation*}
	\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation*}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(\lambda)=e^{-\tau t \lambda}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$, which are a generalization of squared frequencies. Hence, the kernel $K_t$ can also be considered as a generalization of the Gaussian kernel on the sphere. \figref{gaussian_filters_comparizon} shows the effect of the convolution by diffusing a unit of heat for $\tau=1$ at various times $t$. By comparing the graph convolution with the spherical symmetric Gaussian smoothing, we observe that both techniques lead to similar results (see \figref{gaussian_filters_comparizon}). While the graph convolution is different from a true spherical convolution, it can, provided the correct parameterization, well approximate it. \figref{gaussian_filters_visualization} provides additional visualizations of the convolution kernel $K_t$.

@@

This result is another sign that the constructed graph is able to capture the spherical
structure of the HEALPix sampling. In some applications, where the
exactitude of the convolution is not a requirement, such as de-noising, graph convolution could be used instead of using the spherical harmonics.
The fact that both convolutions are not identical is not relevant in a neural network setting, as the convolution kernels are learned anyway.
% not ideal tho as the graph convolution depends on the localization

\subsection{Efficient convolutions}
\label{sec:efficient_convolution}

While \eqnref{graph_convolution} is a theoretically well justified definition of the convolution of a graph signal $\f$ by a kernel $h$, it is computationally cumbersome. As there exists no general fast Fourier transform (FFT) for graphs,\footnote{See \citep{le2018fgft} for a comparison between the polynomial definition of filters presented here and a class of fast graph Fourier transforms based on the approximation of the Fourier basis $\U$ by a product of sparse matrices.} the Fourier transform, i.e., the multiplication of the signal $\f$ by the dense matrix $\U$, costs $\bO(N_{pix}^2)$ operations. In a CNN, this operation has to be performed for each training step (forward and backward). As current training procedures are not data efficient and require a lot of samples, a quadratic computational cost in the sample size is a show stopper.
Moreover, the eigendecomposition of the Laplacian $\L$ is needed to obtain the Fourier basis $\U$ in the first place. That has a unique cost of $\bO(N_{pix}^3)$ operations.

To overcome those computational issues, the convolution kernel $h$ can be defined as a polynomial $h_\theta(\lambda) = \sum_{k=0}^K \theta_k \lambda^k$ of degree $K$ parameterized by $K+1$ coefficients $\theta$. The filtering operation \eqnref{graph_convolution} then becomes
\begin{equation} \label{eqn:graph_convolution_polynomial}
	h_\theta(\L) \f = \sum_{k=0}^K \theta_k \L^k \f.
\end{equation}
Here happens the transition from filtering in the spectral domain to filtering in the vertex domain. Note that $\L^k$ captures $k$-neighborhoods, i.e., $(\L^k)_{ij}$ is non-zero if and only if nodes $v_i$ and $v_j$ are connected by a path of length $k$. As such, filtering with a polynomial convolution kernel can be interpreted in the vertex domain as a local weighted linear combination of the signal's value. The weights are determined by the filter coefficients $\theta$ and the path weight, defined as the multiplication of all edge weights along the path.
\todo{Better vertex domain explanation.}

To make the parallel with classical Euclidean signal-domain convolution, i.e., filtering by multiplying the signal with translated versions of a patch, we here match our filter at the level of neighborhoods\footnote{The 1-neighborhood of a node is the set of nodes which are directly connected to it. The 2-neighborhood is the set of nodes which are connected through paths of length 2.} rather than individual nodes. That is required as there is no notion of ordering in a general graph. For example, in a classical convolution on the grid, the relative ordering of vertices is up, down, left, and right. There is no such ordering on the sphere either, especially if irregularly sampled.

Following \citet{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta(\tL) \f = \sum_{k=0}^K \theta_k T_k(\tL) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_polynomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.

Exploiting the recursive formulation of Chebyshev polynomials, evaluating \eqnref{graph_convolution_cheby} requires $\bO(K)$ multiplications of the dense vector $\f$ with the sparse matrix $\tL$. The cost of one such multiplication is $\bO(|\E|)$. By construction of our graph, $|\E| < 8 N_{pix}$. The overall computational cost of the convolution is thus reduced to $\bO(N_{pix})$ operations.
That is much more efficient than filtering with spherical harmonics, even tho the HEALPix sampling was designed to leverage a fast spherical transform (by being iso-latitude). That is especially true for smooth kernels which require a low polynomial degree $K$. See \figref{filtering_speed} for a speed comparison of low-pass filtering for Gaussian smoothing using the spherical harmonics and the presented method. It is ten to twenty times faster for $N_{side} = 2048$, with $K=20$ and $K=5$, respectively.
